\documentclass{letter}
\usepackage{microtype}
\frenchspacing

\signature{Peter D. Smits}
\address{952 NW 60th St \\ Seattle, WA 98107 \\ peterdavidsmits@gmail.com}
% \address{Department of Integrative Biology \\ University of California Berkeley \\ 3040 Valley Life Sciences Building Rm. 5151 \\ Berkeley, CA 94720 \\ psmits@berkeley.edu}

\begin{document}
\begin{letter}{IHME, University of Washington}
  \opening{Dear Hiring Manager,}

  I am applying for the Research Scientist (Resource Tracking US Spending) position at the Institute for Health Metrics and Evaluation. Based on the description, I think that I am a good candidate for this position because of my background in evolutionary biology as well as my experience as an applied statistician.

Currently, I am a Data Scientist II at Amazon in Advertising. I lead the design and development of multiple data science products using multilevel statistical models to measure the effects of nudges on daily display ad performance. These projects have let me grow my development skill set. I applied tools like Docker to productionize data processing and modeling pipelines on AWS while working with an international team of Software Development Engineers and Business Intelligence Engineers.

Previously, I worked as a Post-Doctoral Researcher in Biology at the University California -- Berkeley were I developed statistical models for describing marine plankton extinction rates over the last 60 million years. This project allowed me to focus on how to communicate uncertainty and out-of-sample prediction error, skill that continue to serve me well. At Berkeley, I also co-taught a graduate level course on statistical methods in paleontology, macroevolution, and macroecology using the R and Stan languages.

Thank you for considering my application. I am looking forward to hearing more about this position.

  \closing{Sincerely,}

  \encl{CV, references}

\end{letter}
\end{document}


